---
title: "Cyclistic: Casual vs Member Ridership"
author: "Gregory Whitmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_notebook: 
    theme: journal
    toc: true
    toc_float: true
      #collapsed: true
    smooth_scroll: true
    toc_depth: 4
    number_sections: true
    css: "html/style.css"
    includes:
      after_body: "html/footer.html"
---

------------------------------------------------------------------------

# `r fa("r-project", fill = "purple")` Environment Setup

```         
readr for loading CSV files
tidyverse for knitting file
conflicted to manage and set conflict repairs
ggplot for visuals
ggsci for visual theme improvement
flexadashboard & fontawesome for final visual layout
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(conflicted)
library(ggplot2)
library(ggsci)
#library(flexdashboard)
library(fontawesome)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

------------------------------------------------------------------------

## `r fa("hard-drive", fill = "purple")` Load Data Set from CSV

```{r dataload, echo=TRUE}
q1_2019 <- read_csv("CSV/Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("CSV/Divvy_Trips_2020_Q1.csv")
```

------------------------------------------------------------------------

## `r fa("broom", fill = "purple")` Clean and Combine

### Verify Column Names

```{r quick-compare}
colnames(q1_2019)
colnames(q1_2020)
```

#### Rename Columns for Consistency

```{r column-rename}
(q1_2019 <- rename(q1_2019
 ,ride_id = trip_id
 ,rideable_type = bikeid
 ,started_at = start_time
 ,ended_at = end_time
 ,start_station_name = from_station_name
 ,start_station_id = from_station_id
 ,end_station_name = to_station_name
 ,end_station_id = to_station_id
 ,member_casual = usertype
 ))
```

### Inspect the dataframes

```{r inspection-quarters}
str(q1_2019)
str(q1_2020)
```

#### Convert ride_id and rideable_type to chr

```{r character-conv}
q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id)
 ,rideable_type = as.character(rideable_type))
```

#### stack quarters data frames into one

```{r stackemup}
all_trips <- bind_rows(q1_2019, q1_2020)#, q3_2019)#, q4_2019, q1_2020)
```

#### remove lat, long, birthyear and gender

```{r removal-unnec}
all_trips <- all_trips %>%
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration"))
```

## Clean and prepare new table "all_trips"

### Load 'all_trips' Preview

```{r quick-compare2}
colnames(all_trips)
#nrow(all_trips)
#dim(all_trips)
#head(all_trips)
#str(all_trips)
summary(all_trips)
```
## `r fa("broom", fill = "purple")` Clean and repair New dataFrame
### Clean member/casual column for consistency
```{r fix-members}
#table(all_trips$member_casual) #view table for memberships
all_trips <- all_trips %>%
 mutate(member_casual = recode(member_casual
 ,"Subscriber" = "member"
 ,"Customer" = "casual"))
table(all_trips$member_casual) #verify table update contains 2 variables
```
### Split Date into columns for better aggregation
```{r splitthedate}
 all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
 all_trips$month <- format(as.Date(all_trips$date), "%m")
 all_trips$day <- format(as.Date(all_trips$date), "%d")
 all_trips$year <- format(as.Date(all_trips$date), "%Y")
 all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
### Add ride_length calculation and make numeric
```{r calc-ridelength}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
# convert ride_length from factor to numeric
#is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length) #verify numeric
```
### Remove the rows of bad data
data frame include a few hundred entries when bikes were removed for service
```{r remove-service}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
## `r fa("magnifying-glass-chart", fill = "purple")` Prepare descriptive analysis
### Analyze ride lengths
```{r ridelengths}
#calculate all ride lengths
mean(all_trips_v2$ride_length) #average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint ride length
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
# or more simply on one line
summary(all_trips_v2$ride_length)
```
### compare casual and members
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
#correct days of week order
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday",
 "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN =mean)
```

