---
title: "Cyclistic: Casual vs Member Ridership"
author: "Gregory Whitmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_notebook: 
    theme: journal
    toc: true
    toc_float: true
    smooth_scroll: true
    toc_depth: 4
    number_sections: true
    css: "html/style.css"
    includes:
      before_body: "html/header.html"
      after_body: "html/footer.html"
---

------------------------------------------------------------------------

# `r fa("r-project", fill = "#296473")` Environment Setup

```         
readr for loading CSV files
scakes for better graph scales
tidyverse for knitting file
conflicted to manage and set conflict repairs
ggplot for visuals
fontawesome for final visual layout
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr) # load csv files
library(scales) # better control of graph scales
library(tidyverse)
library(conflicted) # needed?
library(ggplot2) # 
#library(ggsci)
#library(flexdashboard)
library(fontawesome) # awesome icons in TOC
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

------------------------------------------------------------------------

## `r fa("file-csv", fill = "#296473")` Load Data from CSV

```{r dataload, echo=TRUE}
q1_2019 <- read_csv("CSV/Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("CSV/Divvy_Trips_2020_Q1.csv")
```

------------------------------------------------------------------------

# `r fa("broom", fill = "#296473")` Clean and Combine

## Verify Column Names

```{r quick-compare}
colnames(q1_2019)
colnames(q1_2020)
```

### Rename Columns of 'q1_2019.csv' for Consistency

```{r column-rename}
(q1_2019 <- rename(q1_2019
 ,ride_id = trip_id
 ,rideable_type = bikeid
 ,started_at = start_time
 ,ended_at = end_time
 ,start_station_name = from_station_name
 ,start_station_id = from_station_id
 ,end_station_name = to_station_name
 ,end_station_id = to_station_id
 ,member_casual = usertype
 ))
```

## Inspect the Dataframes

```{r inspection-quarters}
str(q1_2019)
str(q1_2020)
```

### Convert ride_id and rideable_type columns to CHR

```{r character-conv}
q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
```
```{r}
str(q1_2019)
```

# `r fa("broom", fill = "#296473")` Create and Clean New Table 'all_trips'

## Stack Quarter Dataframes into One

```{r stackemup}
all_trips <- bind_rows(q1_2019, q1_2020)
```

## Remove lat, long, birthyear, and gender

```{r removal-unnec}
all_trips <- all_trips %>%
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, tripduration))
```

## Load 'all_trips' and Preview

```{r quick-compare2}
colnames(all_trips)
#nrow(all_trips)
#dim(all_trips)
#head(all_trips)
#str(all_trips)
summary(all_trips) #only one line needed
```

## Clean 'member_casual' Column for Consistency

```{r fix-members}
#table(all_trips$member_casual) #view table for memberships
all_trips <- all_trips %>%
  mutate(member_casual = recode(member_casual ,"Subscriber" = "member" ,"Customer" = "casual"))
table(all_trips$member_casual) #verify table update contains 2 variables
```

## Split Date into Columns for Better Aggregation

```{r splitthedate}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

## Add 'ride_length' Calculation and Make Numeric

```{r calc-ridelength}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
#is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length) #verify numeric
```

## Remove the Rows of Bad Data

data frame include a few hundred entries when bikes were removed for service

```{r remove-service}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

# `r fa("magnifying-glass-chart", fill = "#296473")` Prepare Descriptive Analysis

## Analyze Ride Lengths

```{r ridelengths}
#calculate all ride lengths
#mean(all_trips_v2$ride_length) #average (total ride length / rides)
#median(all_trips_v2$ride_length) #midpoint ride length
#max(all_trips_v2$ride_length) #longest ride
#min(all_trips_v2$ride_length) #shortest ride
# or more simply on one line
summary(all_trips_v2$ride_length)
```

## Compare Casual and Members

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

## Correct Day of Week Order

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN =mean)
```

# `r fa("chart-simple", fill = "#296473")` Analyze Ridership Data by Type and Weekday

```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>% #creates weekday field using wday
    group_by(member_casual, weekday) %>% #groups by usertype and weekday
      summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
        arrange(member_casual, weekday)
```
```{r}

```

## Visualize Number of Rides by Rider Type

```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
    group_by(member_casual, weekday) %>%
      summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
        arrange(member_casual, weekday) %>%
          ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
          scale_y_continuous(labels=comma) +
          geom_col(position = "dodge")
```

## Visualize Average Duration by Rider Type

```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
    group_by(member_casual, weekday) %>%
      summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
        arrange(member_casual, weekday) %>%
          ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
          geom_col(position = "dodge")
```

# `r fa("file-csv", fill = "#296473")` Export Summary File for Further Analysis

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual +
                      all_trips_v2$day_of_week, FUN = mean,)
write.csv(counts, file = 'output_CSV/avg_ride_length.csv')
```

# Insert Tableau Snapshot and URL to Public site for interactive dashboard 